# UK Economic Data {#sec-uk-data}

The United Kingdom has some of the best-documented and most accessible economic data in the world. Four institutions publish the bulk of it, and each has a dedicated R package.

This chapter walks through pulling, cleaning, and combining data from the ONS, Bank of England, HMRC, and OBR — the building blocks of any UK macro analysis.

## The Office for National Statistics

```{r}
#| eval: false
library(ons)
```

The ONS is the UK's national statistical institute. It publishes GDP, inflation (CPI and CPIH), the labour market, wages, trade, retail sales, house prices, productivity, and population estimates.

### GDP

```{r}
#| eval: false
gdp <- ons_gdp()
head(gdp)
```

### Inflation

```{r}
#| eval: false
cpi <- ons_cpi()
```

### The labour market

```{r}
#| eval: false
unemp <- ons_unemployment()
wages <- ons_wages()
```

## The Bank of England

```{r}
#| eval: false
library(boe)
```

The Bank of England publishes monetary policy data: Bank Rate, SONIA (the sterling overnight rate), gilt yields, exchange rates, mortgage rates, consumer credit, and money supply.

### Bank Rate

```{r}
#| eval: false
bank_rate <- boe_bank_rate()
```

### Gilt yields

```{r}
#| eval: false
gilts <- boe_yield_curve(from = "2020-01")
```

## HM Revenue & Customs

```{r}
#| eval: false
library(hmrc)
```

HMRC publishes monthly tax receipt data — a real-time read on economic activity, since tax receipts respond immediately to changes in employment, spending, and corporate profits.

## The Office for Budget Responsibility

```{r}
#| eval: false
library(obr)
```

The OBR produces the UK's official fiscal forecasts. Its Public Finances Databank contains historical fiscal data going back decades.

## Putting it together: a UK macro dataset

A common task is to assemble a single dataset of key UK macro indicators — GDP growth, inflation, unemployment, Bank Rate — for analysis or charting.

```{r}
#| eval: false
library(ons)
library(boe)
library(dplyr)

gdp <- ons_gdp()
cpi <- ons_cpi()
unemp <- ons_unemployment()
bank_rate <- boe_bank_rate()

# Each returns a tidy data frame with a date column.
# Merge on date for a combined dataset.
```

## Exercises

1. Pull UK GDP data with `ons_gdp()` and calculate the quarter-on-quarter growth rate. In which quarter was growth weakest since 2010?

2. Download Bank Rate with `boe_bank_rate()` and CPI inflation with `ons_cpi()`. Plot both on the same chart (hint: you will need to merge on date and use dual axes or facets).

3. Using `hmrc`, pull the latest tax receipts data. Which tax head generates the most revenue?
