# European and International Data {#sec-european-data}

Macroeconomics rarely stops at national borders. Comparing inflation across the euro area, tracking ECB policy alongside the Fed, or benchmarking UK growth against the G7 — all of these require international data.

This chapter covers the ECB Data Portal, the OECD, and the World Bank.

## The European Central Bank

```{r}
#| eval: false
library(readecb)
```

### Policy rates

```{r}
#| eval: false
rates <- ecb_policy_rates(from = "1999-01")
```

### Inflation across the euro area

```{r}
#| eval: false
hicp <- ecb_hicp(c("DE", "FR", "IT", "ES"), from = "2020-01")
```

### Exchange rates

```{r}
#| eval: false
fx <- ecb_exchange_rate(c("USD", "GBP"), from = "2020-01")
```

### Yield curves

```{r}
#| eval: false
yields <- ecb_yield_curve("10Y", from = "2005-01")
```

## The OECD

```{r}
#| eval: false
library(readoecd)
```

The OECD publishes harmonised data across its 38 member countries — making it the natural choice for cross-country comparisons.

## The World Bank

```{r}
#| eval: false
library(WDI)
```

For broader international comparisons — especially involving emerging markets — the World Bank's World Development Indicators cover 200+ countries.

## Exercises

1. Using `readecb`, compare HICP inflation in Germany, France, Italy, and Spain since 2020. Which country had the highest peak inflation?

2. Pull the EUR/USD exchange rate with `ecb_exchange_rate("USD")` and plot it alongside ECB policy rate changes. Do you see a relationship?

3. Using `WDI`, download GDP per capita (PPP) for the G7 countries. Which country has the highest GDP per capita? Has the ranking changed since 2000?
